The sweet spot for local polygenic adaptation: a case study motivated by stickleback

### 2018-09-17 Stickleback_Paper.tex
last of jared edits, rough draft 1

Threespine stickleback fish provide one of the most striking examples of local adaptation. This hemisphere-wide metapopulation includes both marine populations and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
Rapid adaptation occurs when the freshwater haplotype has been maintained,
through historical introgression, in marine environments without tight linkage to 
marine adapted alleles.
%This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)(No)}.
%We do not see evidence for strong effects encouraging genomic clustering of causal alleles
%or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins and maintenance of genetic polymorphism.
A empirical model is the Alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
% \jgg{Bill - more specifics about Bill, Suzie, Kristin  \& or Thom's papers .. ? }
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

\paragraph{New lakes}
To study adaptation in newly appearing freshwater habitats,
we introduce a new set of lakes midway through the simulation.
As with the old lakes, these are a set of discrete demes connected by migration \emph{only} to the marine.
The initial generation of individuals in these new lakes were created as a copy of the marine individuals 
at the generation of introduction.
The new lake's geography acts like an independent copy of the original set of lakes -- in particular,
the two sets of lakes each have 5,000 individuals at all times 
(Gene flow constricted to within the lakes and through migration to the marine environment).
Since this doubles the number of lake-to-marine immigrants,
after this happens the probability that a new marine individual has freshwater parents
is $2m$ instead of $m$.

@@

To determine the ability for the new lakes to select upon the standing variation in the marine at different rates of historical introgression,
We were able compute the trait value of all effect regions in the initial population.
Given the probability that effect region fixes at a given trait value, $2 * x / 45$, where $x$ is the effect size of a haplotype,
we determined the expected total effect sizes of the haplotypes that fix.
Concretely, it is the sum across the 10 effect regions of
$$\sum_{i=1}^N \prod_{j < i} (1 - p_j) p_i x_i$$
where $N$ is the number of genomes per lake, $p_i$ is the probability of fixation of the 
$i^{th}$ haplotype, $x_{i}$ is the effect size of the $i^{th}$ haplotype, and the haplotypes are sorted in decreasing order by $x_{i}$.
These numbers were computed assuming additivity of mutations, meaning the numbers produced are a slight overestimate.

@@

We varied the migration rate, $m$ across separate simulations
from $5 \times 10^{-5}$ to $5 \times 10^{-1}$ by steps one order of magnitude.
From here on out we will refer to these rates of gene flow as running simulations ranging from {0.01 - 100.0}
migrants \emph{per} lake \emph{per} generation.
%i.e., between 0.01 and 100 migrants per generation to and from each of the freshwater lakes.
This had a strong effect on many aspects of adaptation,
including how fast adaptation occurred in each lake,
how many alleles were shared between lakes,
and the population genetic signals left behind.
We find a threshold of gene flow which allows 
the freshwater haplotype to be maintained in the marine without creating linkage to marine alleles. 
below this threshold, selection on pre-existing freshwater alleles to be much less efficient 
due to background selection. \jgg{did I get this right?}

Interestingly, all rates of $m$ aside from the two highest had little affect on populations ability to locally adapt.
All rates of gene flow aside from the lowest, helped both the old lake's and the new lake's freshwater populations share pre-existing freshwater adapted alleles.
The sharing of alleles at $m = 1.0$ resulted in rapid adaptation ($\approx 32$ generations) of the introduced population to adapt to the freshwater environment.
At the lowest migration rate, lakes adapted nearly independently,
while at the highest migration rate, the habitats act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.

@@

\subsubsection*{Origin of introduced freshwater adaptive alleles: a conundrum}
Up to this point, we have quantified rapid and parallel adaptation and found strong evidence that sharing of alleles is (more than likely?) responsible.
However in Figure \ref{fig:MPFAI}A at $0.1$ \emph{mig/gen/lake}, we see sharing of alleles between populations,
yet \emph{Time to adaptation} is still fairly slow ($\approx 2,000$ gens).
This is surprising when considering we see a similar quantity of allele sharing at  $m$ equal to $1.0$ \emph{mig/gen/lake}, 
but much more rapid adaptation ($\approx 30$ gens). 
Initially, this 60X speedup of adaptation might seem to be the result of subsequent migration from the marine after the point of freshwater introduction. 
To investigate this further we have track the origins of all adaptive variants in all new lake extant individuals at the time of adaptation. 
However, in Figure \ref{fig:Origin}, we find that at both $m = 0.1$ and $m = 1.0$ \emph{migs/gen/lake}, the vast majority of adaptive alleles origins were found to propagate 
from individuals in the original generation ("Captured" OR "Pre-Existing").

@@

The alternative explanation is that at a higher migration rate, a higher saturation of marine adapted alleles are linked to freshwater adapted alleles.
In turn, this linkage acts as a barrier rebuilding the freshwater haplotype, in the new lakes, much more difficult as it much wait for recombination to break the links.
With more historical introgression between the two selective pressures, the freshwater haplotype would remain more "intact" and much more efficient to rebuild in the lake.
To confirm this, we decided to look at the trait values of the ten affect regions in each genome of the initial population.
Given the probability that effect region fixes at a given trait value, $2 * x / 45$ where $x$ is the effect size of a haplotype
we were able to determine expected total effect sizes of the haplotypes that fix (detail about computation in methods).
Below, we see the mean of these expected total effect sizes across all lakes.

\begin{center}
\begin{tabular}{ | c | c  | c  | c | }
 \hline
 $m = 0.01$ & $m = 0.1$ & $m = 1.0$ & $m = 10.0$ \\ 
 $0.1227$ &  $4.6976$ & $12.9298$  & $12.2710$\\
 \hline
\end{tabular}
\end{center}

Here, we see a dramatic difference in the total effect sizes of the haplotypes that are likely to fix.
This provides evidence that linkage to marine alleles has a major impact in the introduced population's ability 
to select and rebuild a pre-existing haplotype given the standing variation at the time of introduction. 
\jgg{Below may belong in the discussion / future work}
This also brings to light the role of recombination in a system of migration selection balance. 
On one hand, a lower recombination rate would allow for the freshwater haplotype to remain "intact" 
within the marine environment, however, without significant introgression the haplotype would quickly get selected against in marine populations. 
On the other hand a higher recombination rate would, in theory, allow freshwater adapted alleles more longevity 
as they would hitchhike with marine adapted haplotypes.

@@

\jgg{TODO: Talk about the role recombination plays}

@@

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{semi_final_plots/Haplo_semi_final.png}
  		\caption{
			Phenotypes (\textbf{Top}) (effect size and mutation type) and haplotypes \textbf{(Botton)} as a function of spatial position of the individual 
			possessing the genome. The phenotype of each haplotype is, by definition, the sum of the 
			effect sizes of each mutation in the genome. 
		}
  		\label{fig:Haplo_Pheno}
	\end{center}
\end{figure}

>>>>

Threespine stickleback fish provide one of the most striking examples of local adaptation. This hemisphere-wide metapopulation includes both marine populations and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
Rapid adaptation occurs when the freshwater haplotype has been maintained,
through historical introgression, in marine environments without tight linkage to 
marine adapted alleles.
We find $F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow.
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins and maintenance of genetic polymorphism.
A empirical model is the Alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

\paragraph{New lakes}
To study adaptation in newly appearing freshwater habitats,
we introduce a new set of lakes midway through the simulation.
As with the old lakes, these are a set of discrete demes connected by migration \emph{only} to the marine.
The initial generation of individuals in these new lakes were created as a copy of the marine individuals 
at the generation of introduction.
The new lake's geography acts like an independent copy of the original set of lakes -- in particular,
the two sets of lakes each have 5,000 individuals at all times.
%(Gene flow constricted to within the lakes and through migration to the marine environment).
Since this doubles the number of lake-to-marine immigrants,
after this happens the probability that a new marine individual has parents from old lakes or the new lakes
is $2m$ instead of $m$.

@@

We were also able to use the tree sequence to get information about the individual ``effect regions" 
in all initial genomes of the new lakes at time of introduction.
From this we determined the ability for the new lakes to select upon the standing variation in the marine.
Given the probability that effect region fixes at a given trait value, $2 * x / 45$, where $x$ is the effect size of a haplotype,
we calculated the expected total effect sizes of the haplotypes that fix.
Concretely, it is the sum across the 10 effect regions of
$$\sum_{i=1}^N \prod_{j < i} (1 - p_j) p_i x_i$$
where $N$ is the number of genomes per lake, $p_i$ is the probability of fixation of the 
$i^{th}$ haplotype, $x_{i}$ is the effect size of the $i^{th}$ haplotype, and the haplotypes are sorted in decreasing order by $x_{i}$.
These numbers were computed assuming additivity of mutations, meaning the numbers produced are a slight overestimate.

@@

We varied the migration rate, $m$ across separate simulations
from $5 \times 10^{-5}$ to $5 \times 10^{-1}$ by one order of magnitude steps.
For clarity given our populations sizes of 5,000 diploid individuals, 200 in each lakes,
we will refer to these rates of gene flow as running simulations ranging from {0.01 - 100.0}
migrants \emph{per} lake \emph{per} generation.
The differing parameter values had a strong effect on many aspects of adaptation;
including how fast adaptation occurred in each lake,
how many alleles were shared between lakes,
and the population genetic signals left behind.
We find a threshold of gene flow which allows 
the freshwater haplotype to be maintained in the marine without creating linkage to marine alleles. 
below this threshold, selection on pre-existing freshwater alleles to be much less efficient 
due to background selection. \jgg{did I get this right?}

Interestingly, all rates of $m$ aside from the two highest had little affect on population's ability to locally adapt.
All rates of gene flow aside from the lowest, helped both the old lake's and the new lake's freshwater populations share pre-existing freshwater adapted alleles.
The sharing of alleles at $m = 1.0$ resulted in rapid adaptation ($\approx 32$ generations) of the introduced population to adapt to the freshwater environment.
At the lowest migration rate, lakes adapted nearly independently,
while at the highest migration rate, the habitats act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.

@@

\subsubsection*{Origin of introduced freshwater adaptive alleles}
Up to this point, we have quantified rapid and parallel adaptation and found strong evidence that sharing of alleles is (more than likely?) responsible.
Interestingly, in Figure \ref{fig:MPFAI}A at $0.1$ \emph{mig/gen/lake}, we see sharing of alleles between populations,
yet \emph{Time to adaptation} is still fairly slow ($\approx 2,000$ gens).
This is surprising when considering we see a similar quantity of allele sharing at  $m$ equal to $1.0$ \emph{mig/gen/lake}, 
but much more rapid adaptation ($\approx 30$ gens). 
Initially, this 60X speedup of adaptation might seem to be the result of subsequent migration from the marine after the point of freshwater introduction. 
perhaps a larger influx of adapted alleles are needed in addition to standing variation in the initial generation.
To investigate this further we have track the origins of all adaptive variants in all new lake extant individuals at the time of adaptation. 
Surprisingly in Figure \ref{fig:Origin}, we find that at both $m = 0.1$ and $m = 1.0$ \emph{migs/gen/lake},
the vast majority of adaptive alleles origins were found to propagate from individuals in the original generation ("Captured" OR "Pre-Existing").

@@

The alternative explanation, is that at a higher migration rate, a higher saturation of marine adapted alleles are linked to freshwater adapted alleles.
In turn, the equilibrium of freshwater alleles in marine genomes created by migration-selection balance at lower rates of gene flow acts as a barrier when rebuilding the freshwater haplotype. 
With more gene flow between the two selective pressures, the freshwater haplotype would remain more "intact" and much more efficient to rebuild in the introduced environment.
To confirm this, we decided to look at the trait values of the ten affect regions in each genome of the initial population.
Given the probability that effect region fixes at a given trait value, $2 \times x / 45$ where $x$ is the effect size of a haplotype
we were able to determine expected total effect sizes of the haplotypes that fix (detail about computation in methods).
Below, we see the mean of these expected total effect sizes across all lakes.

\begin{center}
\begin{tabular}{ | c | c  | c  | c | }
 \hline
 $m = 0.01$ & $m = 0.1$ & $m = 1.0$ & $m = 10.0$ \\ 
 $-0.1227$ &  $-4.6976$ & $-12.9298$  & $-12.2710$\\
 \hline
\end{tabular}
\end{center}

Between $m = 0.1$ and $m = 1.0$ we see a large difference in the total effect sizes of the haplotypes that are expected to fix.
Given these numbers were computed assuming additivity (slight overestimate) and individuals are diploid,
at $m = 0.1$ the mean total genotype expected to fix in the populations puts individuals at a trait value of $\approx -8$ 
after selection on alleles haplotypes in the initial population. 
In turn, the population must wait for novel mutation or subsequent migration to provide the alleles for the remaining 2 units before 
attaining a mean average trait value at the optimum (-10).
It follows in Figure \ref{fig:phenotype_ts2}B that the majority time to adaptation is spent within 2 units of mean trait value.
The initial adaptation on selection of standing variation up to that point happens quite rapidly.
This provides evidence that linkage to marine alleles has a major impact in the introduced population's ability 
to select and rebuild a pre-existing haplotype given the standing variation at time of introduction.

@@

@@

\jgg{TODO: Talk about the role recombination plays}
\jgg{Below may belong in the discussion / future work}
This also brings to light the role of recombination in a system of migration selection balance. 
On one hand, a lower recombination rate would allow for the freshwater haplotype to remain "intact" 
within the marine environment, however, without significant introgression the haplotype would quickly get selected against in marine populations. 
On the other hand a higher recombination rate would, in theory, allow freshwater adapted alleles more longevity 
as they would hitchhike with marine adapted haplotypes.

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=1.0\linewidth]{Final_Plots/Haplo_small.pdf}
  		\caption{
			Phenotypes (\textbf{Top}) (effect size and mutation type) and haplotypes \textbf{(Botton)} as a function of spatial position of the individual 
			possessing the genome. The phenotype of each haplotype is, by definition, the sum of the 
			effect sizes of each mutation in the genome. 
		}
  		\label{fig:Haplo_Pheno}
	\end{center}
\end{figure}
